Cybersecurity Impacts of the Covid-19 Pandemic in Italy

### 2021-03-29 article.tex
Added review about sectors' comments

The \textit{Digital Attacks Observatory} (OAD)~\cite{oadweb} is the only independent online survey in Italy about intentional digital attacks on IT systems of companies and public
bodies. It is implemented in collaboration with the Italian Postal and Telecommunication Police.
The last OAD 2020 was the twelfth year of consecutive surveys on cyberattacks in Italy.
The OAD survey is not directed at a predefined set of respondents, but allows potential interested
parties to have full and free access to an online questionnaire, in a totally anonymous way.
The survey has collected responses from 310 enterprises in 2020, in comparison
to 296 and 2019 and 269 in 2018.
Thanks to the number of responses collected and their balanced distribution between companies and public bodies of different size and belonging to different economic
sectors, we can provide now a specific picture of the cyberattacks in Italy.

@@

The OAD 2020 survey
\footnote{The report is written in Italian,
  only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
  (147 A4 pages) and 9 Attachments (39 A4 pages).}~\cite{oad20}
took place during the
Covid-19 pandemic and covered the entire 2019 and the first quarter
of 2020, when the pandemic exploded in Italy. The comparison of data from 2019 to that
from the first quarter of 2020, provided by the same set of respondents,
has statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyberattacks in Italy.
This pandemic has triggered a wide range of cyberattacks, mainly 
caused by the sudden, and partly unexpected, passage to working from home
and to a strong use of every type of IT service on the Internet, as a consequence
of the mobility lockdowns imposed by the Italian authorities. The resulting respondents pool
covers all the production sectors, including public administration,
even if the majority of the respondents companies belong to the ICT sector (30.8\%).
The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account,
as reported by ISTAT~\cite{istat21}, that 99.91\%
of enterprises in Italy are SMEs, with up to 250 employees, and 95\% have fewer than ten employees.
The OAD 2020 survey is therefore able to consider small and 
tiny organizations, that are the vast majority in Italy, and that usually
are not considered in other cybersecurity surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees and, of these, 22.1\% have fewer than ten
employees.

@@

As pointed out in Fig.~\ref{fig:1}, the IT systems considered in the 2020 survey and their cybersecurity are in the medium to high range and the collected information shows a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian IT systems are based on data centers in 
Italy: most of the respondents' companies have medium to small IT systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations related to the European
General Data Protection Regulation (GDPR), have certainly contributed to strengthening cybersecurity measures, and a further improvement 
comes from the growing use of cloud services, where high standard security measures are in place, usually. Organizational measures for cybersecurity, historically missing and neglected
in Italy, have improved in terms of definition of roles and separation of duties, of organizational policies and procedures, and of incident management. These measures often are missing
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly concentrated at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continued training and awareness. A formal and bureaucratic approach to the organizational procedures is still prevalent. Once defined,
such procedures often do not find concrete application and misses periodic updates and operational tests. A significant example is the disaster recovery plans, for 
which, typically, the required alternative IT resources are neither forecast nor allocated. Periodic exercises and simulations of the 
possible disasters are not implemented. Cybersecurity management tools have limited diffusion yet, among the 2020 OAD respondents. This is particularly true for
the most advanced ones, based on artificial intelligence.
Although the use of IoT~\cite{JK21}, industrial automation, robotics systems~\cite{SGLMDXHKCZT21} and systems based on blockchain technology~\cite{RMFF21}
are expanding and can contribute to fight the Covid-19 pandemic.
In the OAD 2020 report, one finds only a few respondents involved in such technologies, which also derives from the limited portion
of their economic sector, interested in such systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the local public administration for territorial control, and so on.
%For these topics, the data that emerged from the survey are considered only as 
%specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.

>>>>

The \textit{Digital Attacks Observatory} (OAD)~\cite{oadweb} is the only independent online survey in Italy about intentional digital attacks on IT systems of companies and public
bodies. It is implemented in collaboration with the Italian Postal and Telecommunication Police.
The last OAD 2020 was the twelfth year of consecutive surveys on cyberattacks in Italy.
The OAD survey is not directed at a predefined set of respondents, but allows potential interested
parties to have full and free access to an online questionnaire, in a totally anonymous way.
The survey has collected responses from 310 enterprises in 2020, in comparison
to 296 and 2019 and 269 in 2018. Thanks to the number of responses collected and their balanced 
distribution between companies and public bodies of different size, belonging to different economic
sectors, and different geographical areas, we can provide now a specific picture
 of the cyberattacks in Italy.

@@

The OAD 2020 survey
\footnote{The report is written in Italian,
  only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
  (147 A4 pages) and 9 Attachments (39 A4 pages).}~\cite{oad20}
took place during the
Covid-19 pandemic and covered the entire 2019 and the first quarter
of 2020, when the pandemic exploded in Italy. The comparison of data from 2019 to that
from the first quarter of 2020, provided by the same set of respondents,
has statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyberattacks in Italy.
This pandemic has triggered a wide range of cyberattacks, mainly 
caused by the sudden, and partly unexpected, passage to working from home
and to a strong use of every type of IT service on the Internet, as a consequence
of the mobility lockdowns imposed by the Italian authorities. The resulting respondents pool
covers all the production sectors (Fig.~\ref{fig:9}), including public administration,
even if the majority of the respondents companies belong to the ICT sector (30.8\%).
However, many of them are represented by low percentages. this reason, not having sufficiently
representative samples, we have not considered particular data analysis and correlations by sectors.
The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account,
as reported by ISTAT~\cite{istat21}, that 99.91\%
of enterprises in Italy are SMEs, with up to 250 employees, and 95\% have fewer than ten employees.
The OAD 2020 survey is therefore able to consider small and 
tiny organizations, that are the vast majority in Italy, and that usually
are not considered in other cybersecurity surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees and, of these, 22.1\% have fewer than ten
employees. The distribution of annual respondents' turnover also reflects the turnover of SMEs, as shown in Fig.~\ref{fig:7}. Regarding geographical distribution (Fig.~\ref{fig:8}), the responders' companies and organizations are located for the most part only on Italian territory (77.9\%), predominantly in northern Italy (43.3\%).

\begin{figure}
	\centering
	\includegraphics[width=0.90\textwidth]{pictures/fig9.png}
	\caption{Product breakdown of the respondents' companies and organizations~\cite{oad20}.}
	\label{fig:9}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.90\textwidth]{pictures/fig7.png}
	\caption{Last annual available turnover of responders' companies and organizations~\cite{oad20}. 
		 It should be noted that almost 20\% of the respondents cannot or do not want provide the turnover class to which they fall, stating that the confidentiality policies prevent to indicating a turnover class, even in an investigation completely anonymous.
	}
	\label{fig:7}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.90\textwidth]{pictures/fig8.png}
	\caption{Geographical coverage of the companies/organizations of respondents~\cite{oad20}.}
	\label{fig:8}
\end{figure}

@@

As pointed out in Fig.~\ref{fig:1}, the IT systems considered in the 2020 survey and their cybersecurity are in the medium to high range and the collected information shows a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian IT systems are based on data centers in 
Italy: most of the respondents' companies have medium to small IT systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations related to the European
General Data Protection Regulation (GDPR), have certainly contributed to strengthening cybersecurity measures, and a further improvement 
comes from the growing use of cloud services, where high standard security measures are in place, usually. Organizational measures for cybersecurity, historically missing and neglected
in Italy, have improved in terms of definition of roles and separation of duties, of organizational policies and procedures, and of incident management. These measures often are missing
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly concentrated at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continued training and awareness. A formal and bureaucratic approach to the organizational procedures is still prevalent. Once defined,
such procedures often do not find concrete application and misses periodic updates and operational tests. A significant example is the disaster recovery plans, for 
which, typically, the required alternative IT resources are neither forecast nor allocated. Periodic exercises and simulations of the 
possible disasters are not implemented. Cybersecurity management tools have limited diffusion yet, among the 2020 OAD respondents. This is particularly true for
the most advanced ones, based on artificial intelligence.
Although the use of IoT~\cite{JK21}, industrial automation, robotics systems~\cite{SGLMDXHKCZT21} and systems based on blockchain technology~\cite{RMFF21}
are expanding and can contribute to fight the Covid-19 pandemic.
In the OAD 2020 report, one finds only a few respondents involved in such technologies, which also derives from the limited portion
of their economic sector, interested in such systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the local public administration for territorial control, and so on.
